---
title: "Subqueries"
author: "Kevin Wang"
date: "8 June 2021"
output: html_document
---

# Data preparation 

Evaluated only once to create a CSV file.

```{r, eval = FALSE}
library(nycflights13)
library(tidyverse)
set.seed(123)
readr::write_csv(x = nycflights13::airports %>% 
                   dplyr::select(faa, name), 
                 file = "data/airports.csv")
readr::write_csv(x = nycflights13::airlines, file = "data/airlines.csv")
readr::write_csv(x = nycflights13::flights %>% 
                   dplyr::slice(1:100) %>% 
                   dplyr::mutate(date = paste(year, month, day, sep = "/") %>% lubridate::ymd()) %>% 
                   dplyr::select(date, carrier, flight, dep_delay, origin, dest), 
                 file = "data/flights_dates.csv")
```

# Setting up connection (in `R`)

This chunk of code will take the in-memory data and allow this data to be queried exactly like a SQL database. 

The print methods for `R` is different to that of `SQL`. 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")
airports = readr::read_csv(file = "data/airports.csv")
airlines = readr::read_csv(file = "data/airlines.csv")
flights = readr::read_csv(file = "data/flights_dates.csv")
dbListTables(con)

dbWriteTable(con, "airports", airports)
dbWriteTable(con, "airlines", airlines)
dbWriteTable(con, "flights", flights)
dbListTables(con)
dbListFields(con, "airports")
dbListFields(con, "airlines")
dbListFields(con, "flights")
```


# `WITH` statement

The example below is a little bit basic. However, I think it illustrate an important difference between `dplyr` and `SQL`. The syntax in `dplyr` is largely consistent and linear, which makes it more human-readable. Whereas `SQL` is a bit more cumbersome. E.g. `WHERE` and `HAVING` operates on records and summary of records respectively when `dplyr` simply uses `filter`. 


::: {.left}
## `R`

```{r}
flights %>% 
  group_by(carrier) %>% 
  summarise(mean_dep_delay = mean(dep_delay)) %>% 
  arrange(desc(mean_dep_delay)) %>% 
  inner_join(airlines, by = "carrier") %>% 
  filter(mean_dep_delay == max(mean_dep_delay))
```
:::

::: {.right}
## `SQL`


<!-- ```{sql, connection=con} -->
<!-- SELECT f.carrier, AVG(f.dep_delay) AS mean_dep_delay -->
<!-- FROM flights AS f -->
<!-- INNER JOIN airlines AS a -->
<!--   ON f.carrier = a.carrier -->
<!-- GROUP BY f.carrier -->
<!-- ``` -->


```{sql, connection=con}
WITH t as (SELECT f.carrier, a.name, AVG(f.dep_delay) AS mean_dep_delay
      FROM flights AS f
      INNER JOIN airlines AS a
        ON f.carrier = a.carrier
      GROUP BY f.carrier)
      
SELECT *
FROM t
WHERE t.mean_dep_delay = (SELECT MAX(t.mean_dep_delay) FROM t);
```

:::

::: {style="margin-bottom:800px;"}
:::
