---
title: "Window"
author: "Kevin Wang"
date: "9 June 2021"
output: html_document
---

# Data preparation 

Evaluated only once to create a CSV file.

```{r, eval = FALSE}
library(nycflights13)
library(tidyverse)
set.seed(123)
readr::write_csv(x = nycflights13::airports %>% 
                   dplyr::select(faa, name), 
                 file = "data/airports.csv")
readr::write_csv(x = nycflights13::airlines, file = "data/airlines.csv")
readr::write_csv(x = nycflights13::flights %>% 
                   dplyr::sample_n(100) %>% 
                   dplyr::mutate(date = paste(year, month, day, sep = "/") %>% lubridate::ymd()) %>% 
                   dplyr::select(date, carrier, flight, dep_delay, origin, dest), 
                 file = "data/flights_dates.csv")
```

# Setting up connection (in `R`)

This chunk of code will take the in-memory data and allow this data to be queried exactly like a SQL database. 

The print methods for `R` is different to that of `SQL`. 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")
flights = readr::read_csv(file = "data/flights_dates.csv")
dbListTables(con)

dbWriteTable(con, "flights", flights)
dbListTables(con)
dbListFields(con, "flights")
```


# Summary over the entire data 

The task is to provide both a summary for the carrier-specific mean departure delay as well as the average across all the data. 

`dplyr` handles this via the use of `group_by`, whereas `SQL` requires `OVER`.

::: {.left}
## `R`

```{r}
flights %>% 
  group_by(carrier) %>% 
  summarise(carrier_mean_delay = mean(dep_delay),
            overall_mean_delay = mean(flights$dep_delay)) %>% 
  ungroup()
```


:::

::: {.right}
## `SQL`

```{sql, connection=con}
SELECT carrier, 
       AVG(dep_delay) AS carrier_mean_delay,
       AVG(dep_delay) OVER() AS overall_mean_delay
FROM flights
GROUP BY carrier
```
:::

::: {style="margin-bottom:800px;"}
:::

# Summary statistics by partitions

::: {.left}
## `R`

```{r}

```
:::

::: {.right}
## `SQL`

```{sql, connection=con}
SELECT carrier, 
  origin,
  AVG(dep_delay) OVER(PARTITION BY carrier) AS carrier_delay,
  AVG(dep_delay) OVER(PARTITION BY origin) AS origin_delay
FROM flights
GROUP BY carrier, origin
```
:::

::: {style="margin-bottom:800px;"}
:::

# Cumulative sum/average

::: {.left}
## `R`

```{r}
flights %>% 
  arrange(date) %>% 
  group_by(carrier, date) %>% 
  dplyr::summarise(running_total_delay = cumsum(dep_delay))
```
:::

::: {.right}
## `SQL`

```{sql, connection=con}
SELECT carrier, 
  date,
  origin,
  SUM(dep_delay) OVER(ORDER BY date ROWS BETWEEN UNBOUNDED PRECEDING AND CURRENT ROW) AS running_total_delay
FROM flights
GROUP BY carrier
ORDER BY carrier, date;
```
:::

::: {style="margin-bottom:800px;"}
:::